`respR` - An R package for the analysis of respirometry data

### 2017-10-16 manuscript/manuscript-HTML.Rmd
manuscript update

Recently, a number of free R software packages designed for respirometry have become available (e.g. `respirometry`: https://CRAN.R-project.org/package=respirometry, `rMR`: https://CRAN.R-project.org/package=rMR, `LoLinR`: Olito et al., 2017). Both `respirometry` and `rMR` are useful for aquatic respirometry, but are not focused on the analysis of data, although both packages have functions for the analysis of metabolic rate over intervals, and `rMR` has a specialised function to calculate $P_{c}$. For `LoLinR`, the package provides a statistically robust way of automatically detecting a best-fit regression which are applicable for closed and flow-through respirometry. However, `LoLinR`'s processing time increases exponentially with large datasets, with data containing more than 500 observations taking a few minutes to run (Olito et al., 2017), which makes `LoLinR` intractable with large amounts of data.

@@

The functions `calc.bg.rate()`, `calc.rate()`, `auto.rate()` and `pcrit()` can process raw respirometry data and are called depending on the type of analysis required.

@@

Background $RO_2$ is measured to account for the influence of bacterial respiration in the chamber. Since background rates typically account for a small percentage of experimental rates, these often-called “blank” experiments are routinely conducted alongside main experiments, and the rates are averaged across several datasets to obtain a more accurate estimate of the correction. The function `calc.bg.rate()` can simultaneously process multiple background rate measurements, as long as they share the time data, and produces an output object that can be used in later functions that accept background rate arguments.

@@

A typical procedure in the analysis respirometry data is to select the most linear section of the data for calculations of $RO_2$ (e.g. citations). The function `calc.rate()` can manually extract one or more segments from a single data frame to process them simultaneously. Data segments can be selected by time period, row numbers, $O_2$ decrease or proportion, which can accomodate most, if not all, data selection requirements and allow for consistent reporting of methods and results (e.g. estimating $RO_2$ over a 10-minute time window across multiple specimens).

Alternatively, we can take advantage of automated methods in `auto.rate()`, which implements novel rolling regression and kernel density estimation algorithms to detect patterns in the data to automatically obtain the maximum, minimum or most linear sections of the data. There are clear advantages in using `auto.rate()` when estimating the prior mentioned metrics. Unlike manual methods where data is most likely arbitrary selected, `auto.rate()` adheres to well-defined rules to obtain the rate of interest, so its methods are results are transparent and fully reproducible. When called, the function takes the data frame of length $m$, and performs a rolling regression of sample size $n$ iteratively across the entire data frame (Fig. 2b). Thus, a total of $(m−n)+1$ number of iterative regressions are fitted, and we obtain a new dataset of regressions as a function of time (Fig. 3), which is then ranked by size to obtain maximum and minimum values.

To estimate the most linear rate, further computations are performed after rolling regressions. Logically, the most linear section of the data should reflect a consistently stable (i.e. flat) value in the rolling regression data. Thus the value will have a high probability density. The function `auto.rate()` takes advantage of this relationship to identify data that contains the most consistent rate over a rolling window (which equate to peaks in the density plot) and then recalculates the rate across the new data window. This is repeated for all linear sections of the data that are identified by the function, and the results are ranked by size.

>>>>

Recently, a number of free R software packages designed for respirometry have become available (e.g. `respirometry`: https://CRAN.R-project.org/package=respirometry, `rMR`: https://CRAN.R-project.org/package=rMR, `LoLinR`: Olito et al., 2017). Both `respirometry` and `rMR` are useful for aquatic respirometry, but are not focused on the analysis of data, although both packages have functions for the analysis of metabolic rate over non-overlapping intervals, and `rMR` has a specialised function to calculate $P_{c}$. For `LoLinR`, the package provides a statistically robust method of detecting a "best-fit" regression which is applicable for closed and flow-through respirometry. However, `LoLinR`'s processing time increases exponentially with large datasets, with data containing more than 500 observations taking a few minutes to run (Olito et al., 2017), which makes `LoLinR` intractable with large amounts of data.

@@

The functions `calc.bg.rate()`, `calc.rate()`, `auto.rate()`, `calc.rate.ft()` and `pcrit()` can process raw respirometry data and are called depending on the experimental setup and type of analysis required.

@@

Background $RO_2$ is measured to account for the influence of bacterial respiration in the chamber, and once measured, is often applied as a correction to the main experiment (Rodgers et al., 2016). Since background rates typically account for a small percentage of experimental rates, these “blank” experiments are routinely conducted alongside main experiments, and the rates are averaged across several datasets to obtain a more accurate estimate of the correction. The function `calc.bg.rate()` can simultaneously process multiple background rate measurements, as long as they share the time data, and produces an output object that can be used in later functions that accept background rate arguments.

@@

A typical procedure in the analysis respirometry data is to select the most linear section of the data for calculations of $RO_2$ (e.g. citations). The function `calc.rate()` can manually extract one or more segments from a single data frame to process them simultaneously. Data segments can be selected by (1) time period, (2) row numbers, (3) $O_2$ decrease or (4) proportion, which should accomodate most, if not all, data selection requirements and allow for consistent reporting of methods and results (e.g. estimating $RO_2$ over a regular 10-minute time window across multiple specimens).

In addition, we can take advantage of automated methods in `auto.rate()`, which implements novel rolling regression and kernel density estimation algorithms to detect patterns in the data to automatically obtain the maximum or minimum rate of fixed width, or to detect the most linear sections of the data. Unlike manual methods where data is most likely arbitrary selected, `auto.rate()` adheres to well-defined rules to obtain the rate of interest, so its methods are results are transparent and fully reproducible. When called, the function takes the data frame of length $m$, and performs a rolling regression of sample size $n$ iteratively across the entire data frame (Fig. 2b). Thus, a total of $(m−n)+1$ number of iterative regressions are fitted, and we obtain a new dataset of regressions as a function of time (Fig. 3), which is then ranked by size to obtain maximum and minimum values.

To estimate the most linear rate, further computations are performed after rolling regressions. Logically, linear sections of the data should reflect a consistently stable (i.e. flat) value in the rolling regression data (Fig. 3b). The function `auto.rate()` takes advantage of this relationship to identify data windows that contain consistent rates over a rolling window, implementing the methods of Sheather & Jones (1991), avaliable in the `stats` package in base R, to select the bandwidth using pilot estimation of derivatives and calculate "peaks" in the density that correspond to stable rate values (Fig. 3c). The function then selects the rolling data windows that fall within the kernel density binwidth, merges all overlapping windows into one, and performs a linear regression on each new window.
